---
title: "**Querying EPA CAMD Facility Attributes Data**"
author: "Robert J. Dellinger"
date: "2025"
output:
  html_document:
    fig_caption: true
    css: default
header-includes:
  # Set page margins using geometry package
  - \usepackage{geometry}
  - \geometry{left=1in, right=1in, top=1in, bottom=1in}
  
  # Enable double-spacing for readability
  - \usepackage{setspace}
  - \doublespacing
  
  # Customize the header and footer using fancyhdr
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead[L]{}
  - \fancyhead[C]{Scientific Hypothesis and Preliminary Data Explorations}
  - \fancyhead[R]{}
  - \fancyfoot[L]{}
  - \fancyfoot[C]{\thepage}
  - \fancyfoot[R]{}
  - \renewcommand{\headrulewidth}{0.2pt}
  - \renewcommand{\footrulewidth}{0pt}
---

------------------------------------------------------------------------

# Overview

This document describes methods for querying facility data from the EPA CAMD API. We first retrieve bulk file metadata, filter for Facility files, download the corresponding files, and combine the data into one dataset. The final aggregated facility data are saved as a CSV file for further analysis.

## Methods

1. **Retrieve Metadata:**  
   - Send a GET request to the bulk files endpoint.
   - Parse the JSON response into a data frame and check available file types.

2. **Filter and Download Facility Files:**  
   - Filter metadata for files with `dataType == "Facility"`.
   - Download each file, clean column names, and combine them into one master data table.
   - Save the aggregated data as `CAMD_facilities_data.csv`.

3. **Summarize Data:**  
   - Compute unique facilities per year (if applicable) to gain insights into facility coverage over time.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load the necessary libraries
library(httr)
library(jsonlite)
library(data.table)
library(janitor)
library(tidyverse)
library(here)
library(leaflet)
```


# Querying Facility Data

This section queries the EPA CAMD API for facility data, specifically focusing on the Facility files. The data is downloaded and saved in CSV format.

```{r facility, echo=TRUE, message=TRUE, warning=FALSE}
# Set your API key here
#api_key <- "YOUR_API_KEY"

# API base URLs
apiUrlBase <- "https://api.epa.gov/easey"
bucketUrlBase <- "https://api.epa.gov/easey/bulk-files/"

# Bulk files endpoint (e.g., CAMD Services)
servicesUrl <- paste0(apiUrlBase, "/camd-services/bulk-files?API_KEY=", api_key)

# GET request to the bulk endpoint
res <- GET(servicesUrl)

# Handle error responses
if (res$status_code > 399) {
  errorFrame <- fromJSON(rawToChar(res$content))
  stop(paste("Error Code:", errorFrame$error$code, errorFrame$error$message))
}

# Convert the JSON response into a data frame
bulkFiles <- fromJSON(rawToChar(res$content))

# Confirm available file types
print(unique(bulkFiles$metadata$dataType))

# Filter for Facility files (corrected label)
facilityFiles <- bulkFiles[bulkFiles$metadata$dataType == "Facility", ]

# Ensure consistent datetime format
bulkFiles$lastUpdated <- strftime(bulkFiles$lastUpdated, "%Y-%m-%dT%H:%M:%S")

# Summary
message("Number of facility files to download: ", nrow(newBulkFiles))
message("Total size of facility files: ", sum(newBulkFiles$megaBytes, na.rm = TRUE), " MB")

# Set download directory
downloadDir <- here("Data", "Raw", "EPA_CAMD_Facilities_Data")
if (!dir.exists(downloadDir)) dir.create(downloadDir, recursive = TRUE)

# Initialize a data table
facilityDataBulk <- data.table()

# Loop through each new file
if (nrow(newBulkFiles) > 0) {
  for (i in 1:nrow(newBulkFiles)) {
    s3Path <- newBulkFiles[i, "s3Path"]
    filename <- newBulkFiles[i, "filename"]
    fileURL <- paste0(bucketUrlBase, s3Path)
    
    message("Downloading file: ", filename, " from URL: ", fileURL)
    
    # Optionally: save to disk
    # GET(fileURL, write_disk(file.path(downloadDir, filename), overwrite = TRUE))

    # Read data directly into R and clean it
    tempData <- fread(fileURL, showProgress = TRUE)
    tempData <- clean_names(tempData)
    
    facilityDataBulk <- rbind(facilityDataBulk, tempData, fill = TRUE)
  }
  
  # Sort by year if column exists
  if ("year" %in% names(facilityDataBulk)) {
    facilityDataBulk <- facilityDataBulk %>% arrange(year)
  }
}

print(facilityDataBulk)


# Save output to CSV
outputPath <- file.path(downloadDir, "CAMD_facilities_data.csv")
write_csv(facilityDataBulk, outputPath)

# Unique facilities per year per state
unique_facilities_year <- facilityDataBulk %>%
    group_by(year) %>%
    summarise(unique_facilities = n_distinct(facility_id), .groups = "drop") %>%
    arrange(year)
print(unique_facilities_year)

```


# Mapping EPA CAMD Facilities

Filter out any records missing latitude or longitude, then create an interactive leaflet map.

```{r map-facilities, message=FALSE}
# Load the facility data CSV.
# Adjust the file path if needed.
facilityData <- read_csv(here::here("Data", "Raw", "EPA_CAMD_Facilities_Data", "CAMD_facilities_data.csv"))

# Inspect the first few rows
head(facilityData)

# Convert commercial_operation_date to Date type if needed
facilityDataBulk <- facilityDataBulk %>% 
  mutate(commercial_operation_date = as.Date(commercial_operation_date))

# Aggregate data by facility (unique facility_id & facility_name)
facility_summary <- facilityDataBulk %>%
  group_by(state, facility_id, facility_name) %>%
  summarise(
    n_units = n_distinct(unit_id), 
    first_operation = min(commercial_operation_date, na.rm = TRUE),
    year_range = paste0(min(year, na.rm = TRUE), "-", max(year, na.rm = TRUE)),
    latitude = first(latitude),
    longitude = first(longitude),
    program_code = first(program_code),
    operating_status = first(operating_status),
    .groups = "drop"
  )

# Inspect summary
print(facility_summary)

# Filter out facilities with missing coordinates
facility_map_data <- facility_summary %>% 
  filter(!is.na(latitude) & !is.na(longitude))

# Define custom awesome markers with FontAwesome icons
customIcon <- awesomeIcons(
  icon = "industry",      # 'industry' icon to represent facilities
  iconColor = "white",
  markerColor = "steelblue",   # Blue markers contrast well on a dark basemap
  library = "fa"
)

leaflet(data = facility_map_data) %>%
  addProviderTiles("CartoDB.DarkMatter") %>%  # Modern dark basemap
  addAwesomeMarkers(
    lng = ~longitude, 
    lat = ~latitude,
    icon = customIcon,
    popup = ~paste0("<strong>", facility_name, "</strong><br>",
                     "Facility ID: ", facility_id, "<br>",
                     "State: ", state, "<br>",
                     "First Operation: ", first_operation, "<br>",
                     "Year Range: ", year_range, "<br>",
                     "Program Code: ", program_code, "<br>",
                     "Units: ", n_units),
    clusterOptions = markerClusterOptions()  # Enable clustering to improve readability
  ) %>%
  addScaleBar(position = "bottomleft") %>%  # Add a scale bar
  addEasyButton(easyButton(
    icon = "fa-globe", title = "Reset View",
    onClick = JS("function(btn, map){ map.setView([39.8283, -98.5795], 4); }")
  ))
```

# Citations

United States Environmental Protection Agency (EPA). “Power Sector Emissions Data.” Washington, DC: Office of Atmospheric Programs, Clean Air Markets Division. Available from EPA’s Air Markets Program Data web site: https://ampd.epa.gov.
