---
title: "**Querying EPA CAMD Emissions Data**"
author: "Robert J. Dellinger"
date: "2025"
output:
  html_document:
    fig_caption: true
    css: default
header-includes:
  # Set page margins using geometry package
  - \usepackage{geometry}
  - \geometry{left=1in, right=1in, top=1in, bottom=1in}
  
  # Enable double-spacing for readability
  - \usepackage{setspace}
  - \doublespacing
  
  # Customize the header and footer using fancyhdr
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead[L]{}
  - \fancyhead[C]{Scientific Hypothesis and Preliminary Data Explorations}
  - \fancyhead[R]{}
  - \fancyfoot[L]{}
  - \fancyfoot[C]{\thepage}
  - \fancyfoot[R]{}
  - \renewcommand{\headrulewidth}{0.2pt}
  - \renewcommand{\footrulewidth}{0pt}
---

------------------------------------------------------------------------

# Overview

This document outlines our method for querying annual emissions data from the EPA CAMD API for each state over the period 1995–2024. We construct a query using pipe-separated values for years and state codes, handle pagination to retrieve all records, and save each state's aggregated data as a CSV file.

## Methods

1. **Query Construction and Metadata Retrieval:**  
   - Build a query with years (1995–2024) and state codes.
   - Use an initial GET request to determine the total number of records from the `x-total-count` header.

2. **Pagination:**  
   - Calculate the total number of pages (with 100 records per page) needed to fetch all data.
   - Loop through pages, appending each page's results until all data are retrieved.

3. **Data Output:**  
   - Combine the pages into a single data frame.
   - Save the combined data for each state to a CSV file.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load the necessary libraries
library(httr)
library(jsonlite)
library(data.table)
library(janitor)
library(tidyverse)
library(here)
```

# Querying Hourly Emissions Data

This section demonstrates how to use the streaming services API to get hourly emissions data. This API is suitable for continuous data streams and avoids the need to page through results.


```{r hourly-emissions, echo=TRUE, message=TRUE, warning=FALSE}

# Set your API key
# api_key <- "YOUR_API_KEY_HERE"
selected_states <- c("CA")

# API info
apiUrlBase <- "https://api.epa.gov/easey"

# streaming hourly emissions endpoint url
hourlyEmissionsUrl <- paste0(apiUrlBase,"/streaming-services/emissions/apportioned/hourly?API_KEY=",api_key)

# api parameters for the streaming emissions/apportioned/hourly endpoint
query <- list(beginDate = "2024-01-01",
              endDate = "2024-01-31",
              stateCode = paste0(selected_states, collapse = '|'))

if (as.Date(query$endDate) - as.Date(query$beginDate) > 31) {
  warning("The request you're making could be too large for the streaming services API and may result in a bad request. Please consider using the bulk data api endpoint instead. An example can be found in the bulk_data_demo.py file.")
}

# making get request using the emissions/apportioned/hourly endpoint
res = GET(hourlyEmissionsUrl, query = query)
# printing the response error message if the response is not successful
if (res$status_code > 399) {
  errorFrame <- fromJSON(rawToChar(res$content))
  stop(paste("Error Code:",errorFrame$error$code,errorFrame$error$message))
}

fieldMappings <- fromJSON(res$headers$`x-field-mappings`)
print(fieldMappings)

# convert response to a data frame
hourlyEmissData <- fromJSON(rawToChar(res$content))

# print head of dataframe
print(head(hourlyEmissData))
```

# Querying Daily Emissions Data

```{r hourly-emissions, echo=TRUE, message=TRUE, warning=FALSE}

# Set your API key
# api_key <- "YOUR_API_KEY_HERE"
selected_states <- c("CA")

# API info
apiUrlBase <- "https://api.epa.gov/easey"

# streaming hourly emissions endpoint url
dailyEmissionsUrl <- paste0(apiUrlBase,"/streaming-services/emissions/apportioned/daily?API_KEY=",api_key)

# api parameters for the streaming emissions/apportioned/daily endpoint
query <- list(beginDate = "2024-01-01",
              endDate = "2024-01-31",
              stateCode = paste0(selected_states, collapse = '|'))

if (as.Date(query$endDate) - as.Date(query$beginDate) > 31) {
  warning("The request you're making could be too large for the streaming services API and may result in a bad request. Please consider using the bulk data api endpoint instead. An example can be found in the bulk_data_demo.py file.")
}

# making get request using the emissions/apportioned/daily endpoint
res = GET(dailyEmissionsUrl, query = query)
# printing the response error message if the response is not successful
if (res$status_code > 399) {
  errorFrame <- fromJSON(rawToChar(res$content))
  stop(paste("Error Code:",errorFrame$error$code,errorFrame$error$message))
}

fieldMappings <- fromJSON(res$headers$`x-field-mappings`)
print(fieldMappings)

# convert response to a data frame
dailyEmissData <- fromJSON(rawToChar(res$content))

# print head of dataframe
print(head(dailyEmissData))

monthlyEmissData <- dailyEmissData %>% 
  mutate(year = year(date),
         month = month(date),
         day = day(date)) %>%
  group_by(facilityName, facilityId, stateCode, year, month) %>%
  summarise(total_so2_mass = sum(so2Mass, na.rm = TRUE),
            total_no2_mass = sum(noxMass, na.rm = TRUE),
            total_co2_mass = sum(co2Mass, na.rm = TRUE))

# print head of dataframe
print(head(monthlyEmissData))
```

# Querying Annual Emissions Data 

Downloading annual emissions data for each state from 1995 to 2024 from the EPA CAMD API. The data includes information on the Acid Rain Program, Cross-State Air Pollution Rule, and Clean Air Interstate Rule. The data is saved in CSV format for each state and data between 1995 and 2024 is included. The code handles pagination to ensure all data is retrieved, and it prints the total number of records for each state. The API key must be set before running the code.

```{r annual-emissions, echo=TRUE, message=TRUE, warning=FALSE}

# Set your API key
# api_key <- "YOUR_API_KEY_HERE"

# API base URLs
apiUrlBase <- "https://api.epa.gov/easey"
# Vector of all U.S. state codes that can be queried (modify if needed)
selected_states <- c("AL", "AR", "AZ", "AR", "AZ", "CA", "CO", "CT", "DC", "DE", "FL", "GA", "IL", "IN", "IA", "ID", "KS", "KY", "LA",
"ME", "MD", "MA", "MI", "MN", "MO", "MS", "MT", "NC", "ND", "NE", "NH", "NM", "NV", "NJ", "NM", "NY", "OK", "OH", "OR", "PA",
"RI", "SC", "SD", "TN", "TX", "UT", "WV", "WI", "WY", "VA", "WA", "WY", "PR")

# downloading all data from the acid rain program, cross state air pollution rule, and the clean air interstate rule
years <- 1995:2024

# Annual emissions endpoint URL
annualEmissionsPageUrl <- paste0(apiUrlBase, "/emissions-mgmt/emissions/apportioned/annual?API_KEY=", api_key)

# Loop through each selected state
for (st in selected_states) {
  message("Processing state: ", st)
  
  # Build query parameters with pipe-separated years
  query <- list(
    year = paste(years, collapse = '|'),
    stateCode = st,
    page = 1,
    perPage = 100
  )
  
  # Make an initial GET request to determine the total number of records
  res <- GET(annualEmissionsPageUrl, query = query)
  if (res$status_code > 399) {
    errorFrame <- fromJSON(rawToChar(res$content))
    stop(paste("Error Code:", errorFrame$error$code, errorFrame$error$message))
  }
  
  totalRowsAvailableForQuery <- as.numeric(res$headers$`x-total-count`)
  message("Total records for ", st, ": ", totalRowsAvailableForQuery)
  
  # Calculate total pages required
  totalPages <- ceiling(totalRowsAvailableForQuery / query$perPage)
  message("Total pages to fetch: ", totalPages)
  
  # Initialize a list to store each page's data
  allData <- list()
  page <- 1
  
  # Loop through pages until all pages are fetched
  while (page <= totalPages) {
    query$page <- page
    res <- GET(annualEmissionsPageUrl, query = query)
    
    if (res$status_code > 399) {
      errorFrame <- fromJSON(rawToChar(res$content))
      stop(paste("Error Code:", errorFrame$error$code, errorFrame$error$message))
    }
    
    dataPage <- fromJSON(rawToChar(res$content))
    # If no rows are returned, break the loop
    if (nrow(dataPage) == 0) break
    
    allData[[page]] <- dataPage
    message("Fetched page ", page)
    page <- page + 1
    
    # After each GET request in your loop
    Sys.sleep(1)  # pause for 1 second
  }
  
  # Combine all pages into a single data frame
  combinedAnnualEmissData <- bind_rows(allData)
  print(combinedAnnualEmissData)
  
  # Write the data to a CSV file for the current state
  outputFile <- here("Data", "Raw", "EPA_CAMD_Annual_Emissions", 
                     paste0("EPA_CAMD_Annual_Emissions_", st, "_1995_2024.csv"))
  write_csv(combinedAnnualEmissData, outputFile)
  message("Saved file: ", outputFile)
}

```


# Citations

United States Environmental Protection Agency (EPA). “Clean Air Markets Program Data.” Washington, DC: Office of Atmospheric Protection, Clean Air Markets Division. Available from EPA’s Air Markets Program Data web site: https://campd.epa.gov/.


